SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-09-14 papers/B1555_EdgeOn/B1555_letter.tex
update for B1555 draft wording
non-smooth, clumpy

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Beware the effect of edge-on discs]{SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios}
\author[Hsueh et al.]{J.-W. Hsueh,$^{1}$\thanks{E-mail: jwhsueh@ucdavis.edu} C. D. Fassnacht,$^{1}$ S. Vegetti,$^{2}$  J. McKean,$^{3,4}$ C. Spingola,$^{4}$ M. W. Auger,$^{5}$
\newauthor  L. V. E. Koopmans,$^{4}$ and D. J. Lagattuta$^{6}$\\
$^{1}$Department of Physics, University of California, Davis, 1 Shields Ave. Davis, CA 95616, USA\\
$^{2}$Max Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, D-85740 Garching, Germany\\
$^{3}$Netherlands Institute for Radio Astronomy (ASTRON), P.O. Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^{4}$Kapteyn Astronomical Institute, University of Groningen, P.O. Box 800, 9700 AV Groningen, The Netherlands\\
$^{5}$Institute of Astronomy, Madingley Road, Cambridge CB3 0HA, UK\\
$^{6}$University of Lyon, 92 Rue Pasteur, 69007 Lyon, France}
\begin{document}

@@

\begin{abstract}
Gravitational lens flux-ratio anomalies provide a powerful technique for measuring dark matter substructure in distant galaxies. However, before using these flux-ratio anomalies to test galaxy formation models, it is imperative to ascertain that the given anomalies are indeed due to the presence of dark matter substructure and not due to some other non-smooth component of the lensing galaxy halo or propagation effects. Here we present the case of CLASS~B1555+375, which has a strong radio-wavelength flux-ratio anomaly, but where the lensing galaxy is also revealed in high resolution infrared imaging with the W. M. Keck-II Telescope and the {\it Hubble Space Telescope} to have a clear edge-on disc component that crosses directly over the pair of images that exhibit the flux-ratio anomaly. We find simple models that include the disc can reproduce the cm-wavelength flux-ratio anomaly without requiring additional dark matter substructure. Although further studies are required, our results suggest the assumption that all flux-ratio anomalies are due to a population of dark matter sub-haloes may be incorrect, and analyses that do not account for the full non-smooth component of the lens macro-model will likely overestimate the substructure mass fraction in massive lensing galaxies.
\end{abstract}

@@

It was first suggested by \citet{Mao1998} that the flux-ratio anomalies observed in the multiple radio-loud images of lensed quasars could be interpreted as a sign of the presence of substructure in lens galaxies. Indeed, small perturbations in the gravitational potential are sufficient to cause flux-ratio anomalies in gravitationally lensed systems \citep{metcalf01,Dalal2002,Bradac02}, particularly for those systems with a close pair of merging images \citep{KD04}. At radio wavelengths, the lensed images are neither sensitive to micro-lensing from stars (\citealt{K03}; although see \citealt{koopmans00}), nor to dust extinction. Also, any variation in the properties of the radio images that are due to propagation effects (e.g. interstellar scattering or free-free absorption) have a well understood frequency dependence that can be identified and accounted for (e.g. \citealt{biggs03,M07,winn04}). This makes the flux-ratio anomalies of radio-loud lensed images a promising tool to detect and measure dark matter substructures in a large sample of distant galaxies \citep{Dalal2002}. However, a recent comparison between the observational data (eight gravitational lens systems with four or more images) and the predictions from a set of cold dark matter simulations that are representative of the mass and redshift of the lens galaxies, suggest that there is only a 1--4 per cent probability that the observed flux-ratio distribution is produced only by substructure \citep{Xu15}.

The \emph{gravitational imaging technique}, which alternatively utilizes the surface brightness distribution of extended arcs and Einstein rings to detect low-mass substructures within lensing haloes, was introduced by \citet{K05} and \citet{V09}. This method has given a measurement of the substructure mass fraction for a sample of 10 lensing galaxies that is currently in agreement with the theoretical expectations from cold matter matter simulations \citep{V14a,V12}. The reason for the conflicting results between the flux-ratio anomaly and gravitational imaging techniques is currently not clear.  While this could be the result of small sample statistics, it could also be an indication that the flux-ratio anomalies have a different origin than substructure \citep[see][for a discussion]{Xu15}.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, as can be seen by comparing the radio contours to the $K^\prime$-band emission  (Fig.~\ref{fig:merlin}), and the faint, but clearly visible emission from what appears to be an edge-on lensing galaxy, with a position angle of $\sim 10$~deg and an ellipticity $e \sim 0.8$, where $e$ is defined in terms of the axis ratio $q as $$q=1-e$.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission that is seen from these images is likely due to dust extinction and additional demagnification due to lensing.

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Beware the effect of edge-on discs]{SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios}
\author[Hsueh et al.]{J.-W. Hsueh,$^{1}$\thanks{E-mail: jwhsueh@ucdavis.edu} C. D. Fassnacht,$^{1}$ S. Vegetti,$^{2}$  J. P. McKean,$^{3,4}$ C. Spingola,$^{4}$ M. W. Auger,$^{5}$
\newauthor  L. V. E. Koopmans,$^{4}$ and D. J. Lagattuta$^{6}$\\
$^{1}$Department of Physics, University of California, Davis, 1 Shields Ave. Davis, CA 95616, USA\\
$^{2}$Max Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, D-85740 Garching, Germany\\
$^{3}$Netherlands Institute for Radio Astronomy (ASTRON), P.O. Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^{4}$Kapteyn Astronomical Institute, University of Groningen, P.O. Box 800, 9700 AV Groningen, The Netherlands\\
$^{5}$Institute of Astronomy, Madingley Road, Cambridge CB3 0HA, UK\\
$^{6}$University of Lyon, 92 Rue Pasteur, 69007 Lyon, France}
\begin{document}

@@

\begin{abstract}
Gravitational lens flux-ratio anomalies provide a powerful technique for measuring dark matter substructure in distant galaxies. However, before using these flux-ratio anomalies to test galaxy formation models, it is imperative to ascertain that the given anomalies are indeed due to the presence of dark matter substructure and not due to some other component of the lensing galaxy halo or to propagation effects. Here we present the case of CLASS~B1555+375, which has a strong radio-wavelength flux-ratio anomaly, but where the lensing galaxy is also revealed in high resolution infrared imaging with the W. M. Keck-II Telescope and the {\it Hubble Space Telescope} to have a clear edge-on disc component that crosses directly over the pair of images that exhibit the flux-ratio anomaly. We find simple models that include the disc can reproduce the cm-wavelength flux-ratio anomaly without requiring additional dark matter substructure. Although further studies are required, our results suggest the assumption that all flux-ratio anomalies are due to a population of dark matter sub-haloes may be incorrect, and analyses that do not account for the full complexity of the lens macro-model will likely overestimate the substructure mass fraction in massive lensing galaxies.
\end{abstract}

@@

It was first suggested by \citet{Mao1998} that the flux-ratio anomalies observed in the multiple radio-loud images of lensed quasars could be interpreted as a sign of the presence of substructure in lens galaxies. Indeed, small perturbations in the gravitational potential are sufficient to cause flux-ratio anomalies in gravitationally lensed systems \citep{metcalf01,Dalal2002,Bradac02}, particularly for those systems with a close pair of merging images \citep{KD04}. At radio wavelengths, the lensed images are neither sensitive to micro-lensing from stars (\citealt{K03}; although see \citealt{koopmans00}), nor to dust extinction. Also, any variation in the properties of the radio images that are due to propagation effects (e.g. interstellar scattering or free-free absorption) have a well understood frequency dependence that can be identified and accounted for (e.g. \citealt{biggs03,M07,winn04}). This makes the flux-ratio anomalies of radio-loud lensed images a promising tool to detect and measure dark matter substructures in a large sample of distant galaxies \citep{Dalal2002}. However, a recent comparison between the observational data (eight gravitational lens systems with four or more images) and the predictions from a set of cold dark matter simulations that are representative of the mass and redshift of the lens galaxies, suggest that there is only a 1--4 per cent probability that the full observed flux-ratio distribution is produced only by substructure \citep{Xu15}.

The \emph{gravitational imaging technique}, which alternatively utilizes the surface brightness distribution of extended arcs and Einstein rings to detect low-mass substructures within lensing haloes, was introduced by \citet{K05} and \citet{V09}. This method has given a measurement of the substructure mass fraction for a sample of 10 lensing galaxies that is currently in agreement with the theoretical expectations from cold matter matter simulations \citep{V14a,V12}. The reason for the conflicting results between the flux-ratio anomaly and gravitational imaging techniques is currently not clear.  While this could be the result of small sample statistics, it could also be an indication that the flux-ratio anomalies have a different origin than clumpy substructure \citep[see][for a discussion]{Xu15}.

@@

There are several notable features in the high-resolution AO and {\it HST} data. These include the nearly complete lack of emission associated with lensed images B and D, as can be seen by comparing the radio contours to the $K^\prime$-band emission  (Fig.~\ref{fig:merlin}), and the faint, but clearly visible emission from what appears to be an edge-on lensing galaxy, with a position angle of $\sim 10$~deg and an ellipticity $e \sim 0.8$, where $e$ is defined in terms of the axis ratio $q$ as $q=1-e$.  We note that the position angle of the lensing galaxy is such that the disc emission lies close to the lensed images B and D. Therefore, the lack of emission that is seen from these images is likely due to dust extinction and additional demagnification due to lensing.
